---
title: "BDA project 2025-2026"
author: "J.A. Westerhuis, A.U.S. Heintz-Buschart"
date: "06/01/2026 - 17/01/2026"
output:
  pdf_document:
    number_sections: no
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

\newpage

# BDA project 2025-2026

## Project description

Welcome to the BDA project 2025-2026 syllabus. In this document you will
find all the information you need to do your BDA project.\

The main goals of the Project are to learn how to :

1.  analyze your data using common data analysis methods,\
2.  visualize the results of the analysis methods,\
3.  be able to interpret the results.\

In this course you can use Quarto or R-Markdown to combine explainable
text and code in the same document. You have to submit a PDF by
rendering of knitting your document. Make sure your PDF only contains
the necessary text and code and not e.g. a print of all your data.

Check the Statistics with R syllabus for tips on working reproducible!

## Exercise description

Besides this project, every topic also has exercises. The goal of the
exercises is to teach you the different properties of the data and of
the data analysis methods:

1.  What are the properties of each sample and how do samples relate to
    each other ?\

2.  What are the properties of the variables, and how are they related
    ?\

3.  What is the distribution of the samples for certain variables ?\

4.  What is the noise level, how much systematic variation is present?\

5.  What is the meaning of scores, loadings, residuals in PCA?\

6.  Which distance measures should be used in clustering methods.\

7.  How to validate the result of a classification method.\

Thus the goal of the exercise is to make you aware of the specifics of
the data and the methods, while the Project is to show you are able to
use R to apply the methods, visualize and interpret results. Therefore,
it is important to first go through the exercises, discuss with your
fellow students, get feedback from TAs and teachers such that you
understand the data and the methods.

NB!

1.  You are responsible for sufficient interaction and feedback.\
2.  In answering the Project questions plagiarism rules still apply,
    thus if you use information other than your own ideas, give proper
    reference.\

### Groups

The Project has to be done in pairs. This is because many of the
open-ended questions we ask during your project are open for debate. In
such cases it is really helpful to have a partner who you can discuss
with about what the next step should be. Also you will be able to run
RStudio on two computers at the same time, which will help you in
splitting up the tasks and completing the assignments faster overall.
Please supply your names in the code block below.

```{r Names, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
name1 = "Andrew Crossley"
name2 = "Cornelus Palsma"
```

### Submission and grading

You will hand in your project quarto or markdown file and knitted .pdf
files in the first and second week. In the first week you will get an
indicative pass/fail grade and some feedback to help you progress. In
the second week you will receive a full grade and some feedback. The
deadline for the first week submission is Tuesday January 13 8:00 (AM),
for which you need to complete Assignments 1 up to and including
Assignment 3. The deadline for the second week submission is Tuesday
January 20 08:00 (AM), for which you will need to complete Assignment 1
up to and including Assignment 6. Note that you can use the week 1
feedback to improve your answers from Assignments 1-3 before the second
submission! If you do, then you have to indicate in the new PDF what was
changed from the previous submission. The latest version of Assignments
1-3 will be used for final grading in week 2. Please refer to the course
book on Canvas for more information on the BDA course organisation.

Grading of the project will be done using a rubric for each assignment.
An overview of the number of points you can get for every assignment is
given below. Every question states how many points you can get for it.

| Assignment                      | Number of points |
|---------------------------------|------------------|
| 1: Linear Algebra               | 20               |
| 2: Gene Expression Analysis     | 20               |
| 3: Principal Component Analysis | 20               |
| 4: Clustering                   | 20               |
| 5: Classification               | 20               |
| 6: ASCA                         | 20               |
| Total                           | 120              |

It is crucial for the graders to understand what you have done and why.
Please elaborate in the supplied text boxes below each question what
your reasoning is for making a certain step. To make your figures
eligible for grading, you should add a clear labels and titles, as well
as a clear description of what you have done to make your figure. When
writing your code in the code blocks, add comments clarifying what you
are doing!

### Root dir

To make sure R knows where to find the data and where to store results
you have to set a root directory. Fill in here the directory on your
computer where the data is stored.

```{r Rootdir, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
knitr::opts_knit$set(root.dir = '..')
```

The following code block contains some extra settings to make sure that
your deliverable at the end of the project will be easier to grade for
the teaching team. Please do not modify it in any way.

```{r truncate output, echo=FALSE, include=FALSE, message=FALSE}
# save the built-in output hook
hook_output <- knitr::knit_hooks$get("output")

# set a new output hook to truncate text output
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n/2), '....', tail(x, n/2 + 1), '\n')
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})
```

### Help

As any programmer will tell you, looking stuff up on Google is a major
part of any project. Hence looking up specific R programming questions
may be helpful in progressing in your project assignments. Coding
websites such as [StackOverflow](https://stackoverflow.com/) are
particularly helpful.

Specific questions on how functions work in R can be answered using the
R documentation. Say you want to understand how you can run the singular
value decomposition function svd(). On the right side of your screen you
can find a "help" tab where you can search for the function that you
want to use. You can type "svd" in the search field there to reach the
documentation of the svd() function. Additionally, you can type "?svd"
in your console window at the bottom of your screen to automatically
search for the documentation of your supplied function.

If things are unclear to you or you have any other question, feel free
to ask any of your lecturers or teaching assistants (TAs):\
- Johan Westerhuis (Course coordinator, teacher
[j.a.westerhuis\@uva.nl](mailto:j.a.westerhuis@uva.nl){.email})\
- Anna Heintz Buschart (Teacher,
[a.u.s.heintzbuschart\@uva.nl](mailto:a.u.s.heintzbuschart@uva.nl){.email})\
- Gustavo Stolf Jeuken (Teacher,
[g.stolfjeuken\@vu.nl](mailto:g.stolfjeuken@vu.nl){.email})\
- Alex Sanchez Cano (Teacher,
[a.sanchezcano\@uva.nl](mailto:a.sanchezcano@uva.nl){.email})\

### Installation of necessary packages

Below are the packages we will use throughout the course. Do not edit
this code block!

```{r load packages, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}

if(!"umap" %in% installed.packages()[,1]){
  install.packages("umap")
}

if(!"MASS" %in% installed.packages()[,1]){
  install.packages("MASS")
}

if(!"gtools" %in% installed.packages()[,1]){
  install.packages("gtools")
}

if(!"stringr" %in% installed.packages()[,1]){
  install.packages("stringr")
}

library(umap)
library(MASS)
library(gtools)
library(stringr)
```

### The MoTrPAC data

The Molecular Transducers of Physical Activity Consortium (MoTrPAC)
profiled the dynamic change of multiple omics layers in several organs
of rats undergoing physical training. The results are presented in the
Nature article "Temporal dynamics of the multi-omic response to
endurance exercise training". The paper can be found on the canvas page,
and the data was obtained from their website MoTrPAC-data.org.

### Experimental design

Six-month-old male and female Fischer 344 rats were subjected to
progressive treadmill endurance exercise training for 1, 2, 4 or 8
weeks. Sex-matched sedentary, untrained rats were used as controls. In
this project we will focus on the transcriptomics data obtained from
blood cells and metabolomics data from plasma. Unfortunately these
measurements do not all come from the same rats.

## Assignments

In this assignment, you will start your exploration of the data by
checking various summaries per row and per column of your data tables.
You will also make a few histograms and inspect them to check the
comparability of the samples and variables. Finally you will mean center
and scale your data for use in the other assignments. NB! We assume you
have worked through the R-tutorial before starting this assignment.

### First look at the data

Load the metabolomics and transcriptomics data sets.

```{r Data importing, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
df_metabolomics = read.csv("Plasma_Metabolomics_Normalized.csv", row.names = 1)
df_transcriptomics = read.csv("Blood_Transcriptomics_Counts.csv",row.names = 1)
```

Take a quick view of the different data sets. Lets start with the
metabolomics data.

```{r metabolomics data example, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
print(df_metabolomics[1:7, c(1:8)])
```

In the first few columns are numbers identifying the specific rat
number, the experimental group (training or control) and experimental
time (1,2,4,8 weeks) as well as the sex of the rat (1 = female, 2 =
male). From column 4 onwards is where the metabolomics data is shown.
For the metabolomics data, the metabolite levels are normalized by the
measurement of a known standard.

Then take a quick view of the transcriptomics data.

```{r transcriptomics data example, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
print(df_transcriptomics[1:7, c(1:8)])
```

The transcriptomics data is transposed, the rats are in the columns
while the genes are in the rows. The experimental details for each rat
can be found in the last few rows of the data frame. Furthermore, the
counts are not considered numeric. Let's change that.

```{r Transpose Transcriptomics data}
df_transcriptomics = data.frame(t(df_transcriptomics))
print(df_transcriptomics[1:7, c(1:8)])
```

### From now on you have to give the code and answers yourself!

### Assignment 1: Linear Algebra (Johan Westerhuis, 05-01-2026)

#### 1A) Data pretreatment:

1)  For the metabolomics data, remove the columns with experimental
    details and Sex of the rat and store them in a new data frame
    df_metabolomics_Exp. Make a new column in the experimental data
    frame called Groups indicating one of these 5 levels: Control 8
    weeks, Training 1 week, Training 2 weeks, Training 4 weeks and
    Training 8 weeks.[1]

2)  Also make a new data frame df_metabolomics_Values with only the
    metabolite levels.[1]

3)  Remove metabolite Oleoyl.EA as this one has missing values for some
    rats. [1]

    !N.B. Do not save the changed file
    "Plasma_Metabolomics_Normalized.csv". At the end of this assignment
    we will save the data under a different name such that
    "Plasma_Metabolomics_Normalized.csv" will not be changed.

```{r Assignment 1A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
df_metabolomics_Exp <- df_metabolomics[, c("Experiment.Group", "Experiment.Time", "Rat.Sex")]
df_metabolomics_Exp$Groups <-paste(df_metabolomics_Exp$Experiment.Group, df_metabolomics_Exp$Experiment.Time, sep = " ")

mgene_cols <- !(colnames(df_metabolomics) %in% c("Experiment.Group", "Experiment.Time", "Rat.Sex"))
df_metabolomics_Values <- df_metabolomics[, mgene_cols]
```

#### 1B)

For the transcriptomics also make data frames with only experimental
details and only with values. Note the experimental details can be found
at the end of the data frame. Also make and add the Group variable to
the df with experimental details [1].

```{r Assignment 1B, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
df_transcriptomics_Exp <- df_transcriptomics[, c("Experiment.Group", "Experiment.Time", "sex")]

df_transcriptomics_Exp$Groups <- paste(
  ifelse(df_transcriptomics_Exp$Training == 0, "Control", "Training"),
  df_transcriptomics_Exp$Time,
  "weeks"
)

gene_cols <- !(colnames(df_transcriptomics) %in% c("Experiment.Group", "Experiment.Time", "sex"))
df_transcriptomics_Values <- df_transcriptomics[, gene_cols]
df_transcriptomics_Values <- data.frame(apply(df_transcriptomics_Values, 2, as.numeric))
rownames(df_transcriptomics_Values) <- rownames(df_transcriptomics)
```

#### To make sure the transcriptomics values are considered numeric, run the next line:

```{r Make transcriptomics counts numeric}
df_transcriptomics_Values = as.data.frame(apply(df_transcriptomics_Values, 2, function(x) as.numeric(as.character(x))))
```

#### 1C)

Many genes only have counts in few rats. Find genes that have nonzero
counts for more than 25 rats. Use the apply function to first calculate
for each gene the number of samples having nonzero counts. Then select
those genes for which this number is larger than 25. Make a new data
frame df_transcriptomics_Filtered that only contains these genes. [1]

```{r Assignment 1C,Filter genes, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
nonzero_counts <- apply(df_transcriptomics_Values, 2, function(x) sum(x != 0))
genes_to_keep <- nonzero_counts > 25
df_transcriptomics_Filtered <- df_transcriptomics_Values[, genes_to_keep]
```

#### 1D)

Now write the metabolomics_Values, metabolomics_Exp,
transcriptomics_Filtered and transcriptomics_Exp data frames into new
csv files in your working directory. [2]

```{r Assignment 1D, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
write.csv(df_metabolomics_Values, "Metabolomics_Values.csv", row.names = TRUE)
write.csv(df_metabolomics_Exp, "Metabolomics_Exp.csv", row.names = TRUE)
write.csv(df_transcriptomics_Filtered, "Transcriptomics_Filtered.csv", row.names = TRUE)
write.csv(df_transcriptomics_Exp, "Transcriptomics_Exp.csv", row.names = TRUE)
```

Check whether the files were written correctly. Now you have cleaned
metabolomics and transcriptomics data which you will explore in the next
assignments.

#### 1E)

1.  Calculate the total sum of the normalized values per row and per
    column of the filtered transcriptome data set using the rowSums()
    and colSums() functions. [1]\
2.  Plot the log of the row and column sums of the filtered
    transcriptomics data in a histogram in a figure with 2 subfigures.
    Add labels for the x-axis and titles for each subfigure. [1]\

```{r Assignment 1E, include=TRUE, eval=TRUE, echo=TRUE}
row_sums <- rowSums(df_transcriptomics_Filtered)
col_sums <- colSums(df_transcriptomics_Filtered)
par(mfrow = c(1, 2))

# Histogram for row sums (rats)
hist(log(row_sums), 
     main = "Log of Row Sums (Rats)", 
     xlab = "Log(Total Counts per Rat)",
     col = "lightblue",
     breaks = 20)

legend("topright", 
       legend = c(
         paste("N:", length(row_sums)),
         paste("Range:", round(min(row_sums)/1e6), "-", round(max(row_sums)/1e6), "M")
       ),
       bty = "n",
       cex = 0.6)

# Histogram for col sums (genes)
hist(log(col_sums), 
     main = "Log of Column Sums (Genes)", 
     xlab = "Log(Total Counts per Gene)",
     col = "lightgreen",
     breaks = 20)

legend("topright", 
       legend = c(
         paste("N:", length(col_sums)),
         paste("Range:", round(min(col_sums)), "-", round(max(col_sums)/1e6, 1), "M")
       ),
       bty = "n",
       cex = 0.7)


par(mfrow = c(1, 1)) 
```

#### 1F): Interpretation.

Write down for each plot what you can learn about the samples and
variables in the data sets.[2]\
Add information of - the meaning of a row sum and of a column sum for
these data. - what does it mean when the sum of all values of a sample
or of a feature is larger than of another sample or feature? - Why is it
useful to plot the log of the row and column sums?

#### Interpretation

Row sums represent total gene expression counts per rat (sequencing
depth or library size), whereas the column sums represent the total
counts per gene across all rats [the overall gene expression level].
Larger row sums indicate higher sequencing depth for a sample, which
typically reflects technical variation rather than biological diffs.
Larger column sums can indicate more highly expressed genes. Log
transformation is essential as the gene data spans many orders of
magnitude and without it the histogram would be dominated by extrem
values. The log values are useful as they compress the scale, make
outliers visible, and help identify the type of distribution the data
best resembles. The row sum histogram is expected to show relatively
consistent sequencing depth across samples, whereas the column sum
histogram is expected to show more variation due to natural genetic
variation in expression levels.

#### 1G)

1.  Calculate the column mean and column standard deviation for the
    filtered transcriptomics data sets using the apply() function. [1]\
2.  Make a scatter plot in which the log(standard deviation) is plotted
    vs the log(mean) of each variable for the transcriptomics
    dataset.[1]\

```{r Assignment 1G, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
col_means <- apply(df_transcriptomics_Filtered, 2, mean)
col_sds <- apply(df_transcriptomics_Filtered, 2, sd)

plot(log(col_means), log(col_sds),
     xlab = "Log(Mean)",
     ylab = "Log(Standard Deviation)",
     main = "Mean-Variance Relationship in Transcriptomics Data",
     pch = 16,
     col = rgb(0, 0, 1, 0.5),
     cex = 0.8)

abline(a = log(1)/2, b = 0.5, col = "red", lty = 2, lwd = 2)

legend("topleft", 
       legend = c(
         paste("N genes:", length(col_means)),
         paste("Mean range:", round(min(col_means)), "-", round(max(col_means))),
         paste("SD range:", round(min(col_sds)), "-", round(max(col_sds)))
       ),
       bty = "n",
       cex = 0.8)

# Add reference line label
text(x = max(log(col_means)) - 1, 
     y = min(log(col_sds)) + 0.5,
     labels = "Poisson reference\n(var = mean)",
     col = "red",
     cex = 0.8)
```

#### 1H): Interpretation.

1.  Explain what is the meaning of the standard deviation of a gene.
    Take into account the knowledge you have about the samples [1].
2.  Explain what information about the data is provided by the
    scatterplot of Exercise 1G. [1]\

#### Interpretation

The standard deviation of a gene measures how much that gene’s
expression level varies across the samples. Since each gene is measured
in multiple samples, the standard deviation quantifies the variability
of expression from sample to sample. A low standard deviation indicates
that the gene is expressed at a similar level across all samples,
suggesting stable or constitutive expression. A high standard deviation
indicates strong variability in expression.

Overall, the plot reveals **heteroskedasticity** in the data:
variability increases with expression level, a characteristic feature of
transcriptomics data that motivates variance-stabilizing transformations
and specialized statistical models.

#### 1I)

Now we go back to the metabolomics data.

1\. Make a new variable that only contains the Alanine levels of the
rats. [0.5]\
2. Meancenter the Alanine levels of the new variable. [0.5]\
3. Scale the meancentered Alanine levels of the new variable with its
standard deviation. [0.5]\
4. Plot the raw, only meancentered and meancentered and scaled Alanine.
levels in a figure with 3 subplots and add the correct labels on the
axes. [0.5]\

#### Put your code here

```{r Assignment 1I, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
alanine_raw <- df_metabolomics_Values$Alanine
alanine_centered <- alanine_raw - mean(alanine_raw)
alanine_scaled <- alanine_centered / sd(alanine_raw)


par(mfrow = c(1, 3))
# Raw Alanine
plot(alanine_raw,
     main = "Raw Alanine Levels",
     xlab = "Sample",
     ylab = "Alanine",
     pch = 16,
     col = "blue")
abline(h = mean(alanine_raw), col = "red", lty = 2)

# Meancentered Alanine
plot(alanine_centered,
     main = "Meancentered Alanine",
     xlab = "Sample",
     ylab = "Alanine (centered)",
     pch = 16,
     col = "darkgreen")
abline(h = 0, col = "red", lty = 2)

# Meancentered and Scaled Alanine
plot(alanine_scaled,
     main = "Centered & Scaled Alanine",
     xlab = "Sample",
     ylab = "Alanine (z-score)",
     pch = 16,
     col = "purple")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

#### 1J): Interpretation.

1.  Explain what centering does to your variable. [1]\
2.  Explain what the combination of centering and scaling does to your
    data. [1]\

#### Interpretation

Centering subtracts the mean of the Alanine levels from each individual
observation. As a result, the new variable has a mean of zero, and all
values represent deviations from the average Alanine level across
samples. Centering shifts the data along the y-axis but doesn't change
the relative distances between observations or the overall spread of the
data.

Centering and scaling (standardization) transforms the data into
z-scores by dividing the centered values by the standard deviation. This
results in a variable with mean zero and standard deviation one. Each
value now represents how many standard deviations it lies above or below
the mean. This makes the data dimensionless and directly comparable to
other metabolites, ensuring that differences in scale do not dominate
multivariate analyses.

#### 1K)

Standard deviations of metabolite levels.

1\. Calculate the standard deviation of each metabolite over all samples
and make a histogram of the standard deviations of the metabolomics
columns, [1].

2\. Discuss what you can learn from the histogram, and how this
information is used with respect to scaling the data. [1]

```{r Assignment 1K, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
metabolite_sds <- apply(df_metabolomics_Values, 2, sd)

hist(metabolite_sds,
     main = "Distribution of Metabolite Standard Deviations",
     xlab = "Standard Deviation",
     ylab = "Frequency",
     col = "lightcoral",
     breaks = 30)
legend("topright",
       legend = c(
         paste("N metabolites:", length(metabolite_sds)),
         paste("Mean SD:", round(mean(metabolite_sds), 2)),
         paste("Median SD:", round(median(metabolite_sds), 2)),
         paste("Range:", round(min(metabolite_sds), 2), "-", round(max(metabolite_sds), 2))
       ),
       bty = "n",
       cex = 0.8)
```

The histogram shows that most metabolites have relatively low and
moderate standard deviations, with some higher variability in the tail
of the right-skewed distribution. This indicates that variability
differs substantially between metabolites. Many metabolites are fairly
stable across samples, whereas a few vary strongly, possibly due to
biological regulation or technical effects.

This information is important for scaling as metabolites with large
standard deviations would dominate downstream analyses simply because of
their larger variance. Scaling ensures that each metabolite contributes
equally to the analysis regardless of its original variability.

### This concludes Assignment 1.

Make sure you save your .Rmd file and workspace so you can continue in
your next project session

### Assignment 2: Gene expression analysis (Douwe Molenaar, 6-01-2026)

In the previous assignment, you explored the expression data of the rats
in terms of counts of mRNA of specific genes. In this assignment we will
look for genes that are different between the Control group and the
Training group at 8 weeks. From the MotrPac website, the
Blood_RNA_Normalized_C8_T8.csv data file was obtained that only consists
of the rats of the training group at 8 weeks and the rats of the control
group. Specific normalization was already applied. In assignment 2 we
will use these data.

Load the normalized datafile Blood_RNA_Normalized_C8_T8.csv from Canvas.

```{r load normalized gene expression data }
df_transcriptomics_Filtered_Norm = read.csv("Blood_RNA_Normalized_C8_T8.csv", row.names = 1)
```

#### 2A)

Transpose the data and remove the sex variable.[1]

```{r Assignment 2A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
df_transposed = as.data.frame(t(df_transcriptomics_Filtered_Norm))
df_transposed$sex = NULL
```

#### 2B)

1.  Check whether the data is homoscedastic by plotting the log of the
    std vs the log of the mean per group.[1]
2.  Make indexes for the control group and the 8week training group[0.5]
3.  Calculate the mean and std per group for each gene.[0.5]
4.  Plot the std vs the mean in a loglog plot.[1]

```{r Assignment 2B, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
#1
mean = apply(df_transposed, 2, mean)
std = apply(df_transposed, 2, sd)

plot(log10(mean), log10(std), col = "blue", pch = 16,
     xlab = "log10(mean)", ylab = "log10(std)", 
     main = "Homoscedasticity Check: log(std) vs log(mean)")

#2
control_idx = grep("^Control", rownames(df_transposed))
training_idx = grep("^Training", rownames(df_transposed))
#3
mean_control = apply(df_transposed[control_idx, ], 2, mean)
std_control = apply(df_transposed[control_idx, ], 2, sd)
mean_training = apply(df_transposed[training_idx, ], 2, mean)
std_training = apply(df_transposed[training_idx, ], 2, sd)
#4
plot(log10(mean_control), log10(std_control), col = "blue", pch = 16,
     xlab = "log10(mean)", ylab = "log10(std)", 
     main = "Homoscedasticity Check: log(std) vs log(mean)")
points(log10(mean_training), log10(std_training), col = "red", pch = 16)
legend("topleft", legend = c("Control", "Training 8 weeks"), 
       col = c("blue", "red"), pch = 16)
```

#### 2C: Interpretation

1.  Explain what kind of figure you would have expected for
    homoscedastic data.[1]
2.  Conclude about the homoscedasticity of this data based on the figure
    you obtained.[1]

#### Interpretation

For homoscedastic data, we would expect to see points scattered randomly
around a horizontal line in the log(std) vs log(mean) plot, with no
clear pattern or trend. The standard deviation would remain relatively
constant across all mean values, meaning the variance does not depend on
the mean. The points would show no systematic increase or decrease in
log(std) as log(mean) changes. The data is heteroscedastic (not
homoscedastic). The plot shows a mildly negative linear relationship
between log(std) and log(mean), indicating that genes with higher mean
expression have proportionally lower standard deviation. This means the
variance decreases with the mean expression level. This
heteroscedasticity violates the assumption of constant variance required
for many statistical tests. \#### End of text block

In the next assignments we will apply tests to find genes related to
training by comparing the control group with the group of rats that
trained 8 weeks.

#### 2D)

First we filter Genes for which there is almost no difference between
the samples (standard deviation \< 0.1) as these will provide problems
in the t-test. This can occur when e.g. for one of the group all rats
had zero counts. 1. Calculate the standard deviation for each gene over
all samples.[1] 2. Only keep those genes with a standard deviation
larger than 0.1.[1]

```{r Assignment 2D , include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
gene_sd = apply(df_transposed, 2, sd)
df_filtered = df_transposed[, gene_sd > 0.1]
cat("Original number of genes:", ncol(df_transposed), "\n")
cat("Filtered number of genes:", ncol(df_filtered), "\n")
```

#### 2E)

Perform a t-test of the expression of Gene ENSRNOG00000000024 between
the Control and Training 8 weeks conditions. [1]\

```{r Assignment 2E , include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
gene_name = "ENSRNOG00000000024"
control_expr = df_filtered[control_idx, gene_name]
training_expr = df_filtered[training_idx, gene_name]
t_test_result = t.test(control_expr, training_expr)
print(t_test_result)
```

#### 2F) Interpretation.

1.  What is the hypothesis being tested? [1]\
2.  What does this p-value mean? [0.5]\
3.  What is your conclusion from the test result? [0.5]\

#### Put any text for Interpretation 2F here.

The null hypothesis is that there is no difference in mean expression of
gene ENSRNOG00000000024 between Control and Training groups. The p-value
represents the probability of observing a difference as extreme as (or
more extreme than) what we observed, assuming the null hypothesis is
true. With a p-value of 0.01292, we reject the null hypothesis and
conclude there is a significant difference in expression between groups.
\#### End of text block.

#### 2G)

1.  Perform a t-test for every gene in the gene expression dataset for
    Control and 8 weeks training using the apply() function, and store
    the p-value output of the t-test in a vector. [0.5]\
2.  Which gene has the smallest p-value? [0.5]

```{r Assignment 2G, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
p_values = apply(df_filtered, 2, function(gene_expr) {
  t_test = t.test(gene_expr[control_idx], gene_expr[training_idx])
  return(t_test$p.value)
})

# Find gene with smallest p-value
min_p_gene = names(which.min(p_values))
cat("\nGene with smallest p-value:", min_p_gene, "\n")
cat("P-value:", min(p_values), "\n")

```

#### Put any text for Interpretation 2G here.

The gene with the smallest p-value is identified as "ENSRNOG00000005576"
which has a p-value of 3.029946e-06. \#### End of text block

#### 2H)

1.  Make a histogram of all the p-values. Make a rough guess of how many
    positives are present in your set of genes? [1]

<!-- -->

2)  How many false positives do you expect given the p-value histogram?
    [1]

```{r Assignment 2H, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
hist(p_values, breaks = 50, col = "lightblue", border = "white",
     xlab = "P-value", main = "Distribution of P-values from T-tests")
n_tests = length(p_values)
n_sig = sum(p_values < 0.05)
expected_false_pos = n_tests * 0.05

cat("\nTotal tests:", n_tests, "\n")
cat("Significant at p < 0.05:", n_sig, "\n")
cat("Expected false positives (5% of", n_tests, "):", expected_false_pos, "\n")
cat("Estimated true positives:", n_sig - expected_false_pos, "\n")
```

#### Put any text for interpretation 2H here.

Over 13900 total tests, 2748 were significant at a p-value less than
0.05. We calculate the expected false positives by finding the excess
above the 5% expected true positives. The Expected false positives are
5% of the 13900, which is 695, therefore the estimated true positives
are 2089. \#### End of text block.

Our gene expression data is already normalized and log transformed, so
we can instead calculate the fold change using the averages between the
two conditions.\

#### 2I)

Calculate the fold change of all genes between Control and 8 weeks
training conditions this way. [1]

```{r Assignment 2I, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
mean_control_all = apply(df_filtered[control_idx, ], 2, mean)
mean_training_all = apply(df_filtered[training_idx, ], 2, mean)
fold_change = mean_training_all - mean_control_all
```

As you may be aware, you have now done almost 14.000 t-tests. As such,
you will need to control for Type I error. In the exercises you did
before the project part of today's computer practical you have
calculated the FDR value for any p-value and plotted them. An easy
function to obtain the FDR values is p.adjust().

#### 2J)

1.  Calculate the adjusted p-values using the p-adjust() function. [1]\
2.  Make a volcano plot using your Fold change and adjusted p-values for
    every gene in a similar way as was done in the exercises of this
    topic. [1]\
3.  Set a threshold for both the adjusted p-value and for the Fold
    Change, add lines to the plot indicating the threshold values and
    show the genes that are differentially expressed according to your
    thresholds in a different color. [1]\

```{r Assignment 2J, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# Calculate adjusted p-values
adj_p_values = p.adjust(p_values, method = "fdr")
p_threshold = 0.05
fc_threshold = 0.5  # Adjust based on your data
sig_genes = (adj_p_values < p_threshold) & (abs(fold_change) > fc_threshold)

# Make Volcano plot
plot(-log10(adj_p_values) ~ fold_change,
     col = ifelse(sig_genes, "red", "gray"),
     pch = 16, cex = 0.8,
     xlab = "Fold Change (log2)",
     ylab = "-log10(Adjusted P-value)",
     main = "Volcano Plot: Control vs Training (8 weeks)")

# set thresholds and find differentially expressed genes
abline(h = -log10(p_threshold), col = "blue", lty = 2, lwd = 2)
abline(v = c(-fc_threshold, fc_threshold), col = "blue", lty = 2, lwd = 2)
legend("topright", 
       legend = c("Significant", "Not significant"), 
       col = c("red", "gray"), pch = 16)
cat("\nNumber of differentially expressed genes:", sum(sig_genes), "\n")

# Most significant gene
most_sig_gene = names(which.min(adj_p_values))
cat("Most significant gene:", most_sig_gene, "\n")
cat("Adjusted p-value:", min(adj_p_values), "\n")
cat("Fold change:", fold_change[most_sig_gene], "\n")
```

#### Put any text you may want to type here.

#### End of text block.

#### 2K) Interpretation.

1.  Explain what it means to control the type I error of your p-values
    using a false discovery rate. [1]\
2.  Find the name of your most important gene and describe its function.
    [1]\

#### Put any text for interpretation 2k here.

False Discovery Rate (FDR) control: When performing multiple tests, the
FDR controls the expected proportion of false positives among all
significant results. FDR is appropriate for exploratory analyses like
gene expression studies where we expect many true positives. For
example, an FDR-adjusted p-value \< 0.05 means we expect less than 5% of
our significant results to be false positives. The most important gene
was found to be "ENSRNOG00000004487" with an adjusted p-value of
0.01565575 and a fold change of 1.181462. According to GeneGlobe the
gene is *"Predicted to enable protein kinase activity and protein
phosphatase binding activity. Predicted to be involved in centrosome
separation; protein autophosphorylation; and regulation of cell cycle
process. Predicted to act upstream of or within several processes,
including blastocyst development; mitotic spindle assembly; and positive
regulation of telomere maintenance. Predicted to be located in several
cellular components, including condensed chromosome; microtubule
cytoskeleton; and midbody."*
\*[source](https://geneglobe.qiagen.com/us/knowledge/gene/ENSRNOG00000004487)

#### End of text block.

#### This concludes Assignment 2.

Make sure you save your .Rmd file and workspace so you can continue in
your next project session.

### Assignment 3: Principal Component Analysis (Gustavo Stolf Jeuken, 8-01-2026)

During this Assignment, you will perform Principal Component Analysis on
both the gene expression and metabolomics data containing all
conditions. You will investigate groupings in the score plots, and have
a look at important variables distinguishing male and female rats, as
well as features that can distinguish the different training durations.

#### 3A)

0.  Load the Metabolomics_Exp and Metabolomics_Values data again and
    meancenter your data.
1.  Perform a principal component analysis on the meancentered
    metabolomics data using the svd() function. [1]\
2.  Make a scatter plot of the scores using the first and second
    principal component. Provide the scores of each treatment group in a
    different color. [1]\
3.  Calculate the variance explained for each component. [0.5]
4.  Show the variance explained of each component in your plot labels.
    [0.5]\
5.  Add a legend to identify the colour for each condition using the
    legend() function. [0.5]\
6.  Also make another scoreplot, similar to the first one, but now plot
    the different sexes in different colors [0.5].

```{r Assignment 3A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
#0 

#1 

#2

#3

#4

#5

#6

```

#### Put any text you may want to type here.

#### End of text block.

#### 3B: Interpretation.

1.  What is your interpretation? Which source of variation is dominating
    the metabolomics data [1]? Include the percentage of variation
    explained per component in your answer.
2.  Discuss whether there is an effect of training in the metabolomics
    data, and if so, how it can be observed. [1]

#### Put any text for Interpretation 3B here.

#### End of text block.

#### 3C: Loading plot.

The loadings of the PCA link the variation found in between the rats to
the metabolites being measured. To circumvent a very crowded loading
plot, only plot the metabolite names of the 30 most important loadings.
$Importance_j =\sqrt(p_{1j}^2 + p_{2j}^2)$. 1: Calculate importance of
each metabolite.[1] 2: Select the 30 metabolites with highest
importance.[2] 1. Make a scatter plot with the names of the important
metabolites.[1] 2. Add again the percentage explained per PC along the
axes.

```{r Assignment 3C Loading plot, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

#1 

#2

#3

```

#### 3D) Interpretation

1.  From the loading plot select some metabolites that are affected by
    training. Discuss why these metabolites could be affected [1].
2.  From the loading plot select some metabolites that are affected by
    Sex. Discuss why these metabolites could be affected.[1]

#### Name some metabolites and give an explanation

#### End of text block.

#### 3E)

1.  Reconstruct your metabolomics dataset using only the first and
    second PCs. [1]\
2.  Calculate the residual matrix of the metabolomics data by
    subtracting the reconstructed data from the original data. [0.5]\
3.  Calculate the Sum of squares of the residuals by first squaring all
    values in the residual matrix and then summing all these values [1]\
4.  Show that the explained variation and the residual variation add up
    to the total variation of the original data [.5]\

```{r Assignment 3E, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### 3F)

1.  Calculate the sum of of the squared residuals over all rows and make
    a bar chart of these sums. Indicate the female rats and male rats
    with different color to show whether there is a Sex difference in
    residual variation, and add a legend. [1]\
2.  Calculate the sum of of the squared residuals over all columns and
    make a bar chart of these sums.[1]\

```{r Assignment 3F, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Put any text you may want to type here.

#### End of text block.

#### 3G

1.  Which rat has the highest residual variation? [1]

2.  Which variable has the highest contribution to the residuals for
    this rat? [1]\

3.  If we define that the variable with the highest % explained
    variation is the best modeled variable, then which variable has been
    modeled the best in the PCA model with 2 PCs? [1]\

#### Put any text you may want to type here.

```{r 3G, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### That concludes week 1 of the project!

Please hand in your .rmd and pdf files on Canvas for your pass/fail
grade and feedback. The week 1 deadline is Tuesday January 13, 8:00 (AM)

### Assignment 4: Clustering (Gustavo Stolf Jeuken, 13-01-2026)

In this assignment, you will focus on clustering your metabolomics data.
This will be done using hierarchical clustering, kmeans and UMAP
methods.\

In the code block below, some functions are defined for you to use in
subsequent questions. You do not need to understand what happens in this
code. Do not edit the code block!

#### 4A)

Import the metabolomics_Values and metabolomics_Exp data again using the
read.csv() function.

```{r Assignment 4A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### 4B)

1.  Meancenter and scale the metabolomics data before clustering using
    the scale function. [1]
2.  Calculate the Euclidean distances between the samples in the
    centered and scaled metabolomics expression data using the dist
    function. [0.5]\
3.  Use the hclust function to calculate the hierarchical clustering
    from the Euclidean distance matrix using average linkage.[0.5]\
4.  Plot the hierarchical clustering dendrogram with labels indicating
    Sex [0.5]
5.  5\. Plot the hierarchical clustering dendrogram with labels
    indicating Group [0.5]

```{r Assignment 4B, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Put any text you may want to type here.

#### End of text block.

#### 4C) Interpretation.

1.  Discuss your interpretation of the cluster result. Focus on how well
    the Rats with respect to Sex and Groups are clustered. [1]\
2.  Discuss the differences between the clustering and the PCA results.
    [1]\

#### Put any text for interpretation 4C here.

#### End of text block.

#### 4D)

Now perform a hierarchical clustering of the METABOLITES of the
meancentered and scaled version of the metabolomics data using Euclidean
distance and average linkage and plot the dendrogram using the
plot_hclust function. To keep the plots visible only apply this on the
first 36 metabolites, which are amines. [2]\

```{r Assignment 4D, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Put any text you may want to type here.

#### End of text block.

#### 4E)

1.  Select 2 different clusters of two metabolites (e.g. Isoleucine and
    Leucine, and Ethanolamine and Hydroxylysine) and plot their
    metabolites levels in a scatter plot to explore their similarity.[2]
2.  Discuss the difference in interpretation when clustering samples
    (experimental conditions) or clustering variables (metabolites). [1]

```{r Assignment 4E}

```

#### Put any text for interpretation of 4E2 here.

#### End of text block.

#### 4F)

1.  Perform a kmeans clustering with 2 clusters and 5 clusters of the
    samples of the metabolomics. [1]\
2.  Make a table of clusters vs experimental groups for both
    clusterings. [1]
3.  Apply kmeans to cluster metabolites. Only select the first 36
    metabolites which are the Amines. Vary the number of clusters based
    on the results in 4D. Show which Amines are grouped in the
    clusters.[1]\

```{r Assignment 4F, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Put any text you may want to type here.

#### End of text block.

#### 4G) Interpretation.

1.  Summarize the kmeans result. Which experiments are clustered
    together? [1]\
2.  Compare the kmeans cluster result with the PCA result. Discuss
    differences and similarities. [1]\
3.  Compare the hierarchical cluster result of the samples with the
    kmeans result. Discuss differences and similarities. [1]\

#### Put any text you may want to type here.

#### End of text block.

#### 4H) UMAP.

1.  Apply a UMAP on the samples of the centered and scaled metabolomics
    data. Vary the number of neighbours until you have a "good" result.
    [0.5]\
2.  Apply a UMAP on the metabolites. For simple visualization reasons
    only use the first 36 metabolites, which are amines. Vary the number
    of neighbours until you have a "good" result.[0.5]

Note! Fix the random state such that the same can be repeated!

```{r Assignment 4H, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Put any text you may want to type here.

#### End of text block.

#### 4I) Interpretation.

1.) What is your interpretation of the UMAP clusters of the metabolite
samples.[1]\
2.) What is your interpretation of the UMAP clusters of the metabolites
in comparison with kmeans and hierarchical clustering.[1]\
3.) Discuss the differences between UMAP and k-means or hierarchical
clustering. [1]\

#### Put any text you may want to type here.

#### End of text block.

#### This concludes Assignment 4.

Make sure you save your .Rmd file and workspace so you can continue in
your next project session.

### Assignment 5: Classification (Johan Westerhuis, 15-01-2023)

In this assignment we will use the LDA and PCDA methods to find the
metabolites that best discriminate between Control and 8 weeks training
conditions.\

#### 5A)

0)  Load the df_metabolomics_Values and the df_metabolomics_Exp
    datasets.
1)  Make a reduced metabolomics dataset by selecting only the samples of
    Training 8 weeks and Control 8 weeks. [1]\
2)  For visualization purposes, only select metabolites 80 (CAR(10:0-OH)
    until metabolite 613 (TG 60:4). These are lipids obtained with
    reversed phase liquid chromatography positive mode.[1]
3)  Center and scale this data set again. [1]\
4)  Create a vector of binary values encoding the condition of each
    sample (0 = Control 8 weeks and 1 = Training 8 weeks). [1]\

```{r Assignment 5A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Put any text you may want to type here.

#### End of text block.

#### 5B)

Perform a linear discriminant analysis on your new metabolomics dataset,
using your new vector of binary values as the class of each sample. [2]

```{r Assignment 5B, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Put any text you may want to type here.

#### End of text block.

#### 5C) Interpretation.

You should have gotten a warning from your LDA function. Explain this
warning. [2]\

#### Put any text for Interpretation 5c here.

#### End of text block.

#### 5D)

To solve the problem being discussed in 5C, you will make a PCDA model
instead of an LDA model.

1\. Apply Principal Component Analysis using the svd function and select
three components. [2]\
2. Make scoreplots of PC1 vs PC2 and PC2 vs PC3 in which the samples are
colored based on their group. [1]\
3. Which of the PCs seems most useful for the discrimination between the
two groups? Think of a way to quantify this. [2]

```{r Assignment 5D, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# PCA

# scoreplot PC1 vs PC2

# scoreplot PC2 vs PC3

#3

```

#### 5E)

1.  Make a PCDA model using the PCA scores and the vector of binary
    class labels .[1]\
2.  Calculate the PCDA regression coefficients of each variable.[1]\
3.  What are the names of the 5 most important metabolites based on
    their absolute PCDA regression coefficients. [1]\
4.  Use the PCDA regression coefficients to make a prediction for each
    rat whether they belong to the control or training group.[1] 5. How
    many misclassifications did you make? [1]

```{r Assignment 5E, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### How many misclassifications did you make?

#### End of text block.

#### 5F) Interpretation.

1.  Discuss the similarities and differences between the PCA and PCDA.
    [1]\
2.  Discuss the difference between using t-test to make a Volcano plot
    and using PCDA to calculate important metabolites [1].

#### Put any text for interpretation 5F here.

#### End of text block.

### Assignment 6: ASCA (Johan Westerhuis, 16-01-2026)

For ASCA we will use the HDANOVA and the mixlm packages. Obtain the
latest version from CRAN.

```{r}
install.packages("mixlm")
install.packages("HDANOVA") 

library(mixlm)
library(HDANOVA)
```

Load the metabolomics_Values and the metabolomics_Exp files again on
which the ASCA model will be applied. Use the same set of features as in
the Classification assignment.

#### 6A)

Apply preprocessing of your data. Consider transformation, centering
and/or scaling the data. Make plots of the data (before and after the
preprocessing steps) to defend your choices. [3]

#### Add code for figures here

```{r Assignment 6A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Add arguments here

#### 6B) Data handling

Make the data from wide to long format that is necessary for the asca
function in the HDANOVA package. Make sure the factors Sex and Group are
defined as factors. (See ASCA exercise) [2]

```{r Assignment 6B, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}


```

#### 6C) ASCA model:

1)  Create an ASCA model using the only the main factors Sex and Group.
    Do not consider interaction terms yet. [1]
2)  Explain the meaning of Sum.Sq and Expl.var. from the ASCA table, and
    explain how they are calculated. [1]
3)  What is your interpretation of the ASCA model? [1]

```{r Assignment 6C, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Give explanation and interpretation here

#### End of interpretation

#### 6D)

1.  Show score and loading plots of PC1 vs PC2 of the Group factor.
    [0.5]
2.  Show the loading plots of PC1 vs PC2 of the Group factor. Only show
    the 20 most important variables where
    $Importance_j =\sqrt(p_{1j}^2 + p_{2j}^2)$. [1.5]
3.  Give your interpretation of the score and loading plots. [1]

```{r Assignment 6D, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Give interpretation of 6D3 here

#### 6E) Add the two-factor interaction Sex:Group to your model.

1.  Make a new model in which the two-factor interaction Sex:Group is
    also included in the model.[1]
2.  What is your conclusion of this ASCA model with interaction? [1]

```{r Assignment 6E ASCA permutation analysis, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Give your conclusion of the interaction model

#### End of conclusion

#### 6F: Statistical Inference

1.  To conclude whether the main factors and interaction are
    significant, apply a permutation test and plot the permutation
    result using the permutation plot function for both main factors and
    the interaction.[1]
2.  Explain why in ASCA a permutation test is used and not an F-test.[1]
3.  What is your interpretation of the permutation test? [1]

```{r Assignment 6F ASCA permutation analysis, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Give explanation of 6F2 here

#### Give interpretation of 6F3 permutation test here

#### 6G)

When exploring the model with interaction, it is common to combine the
interaction effect with one of the main factor effects to improve
interpretation.

1\. Make an ASCA model in which the Sex:Group interaction effect is
combined with the main Group effect.[1]

2.  Make two score plot (using the timeplot function) of the combination
    effect. In the first the PC1 scores of the Females and Males are
    plotted as a function of the Groups. In the second the PC1 scores of
    the Groups are plotted as a function of the Females and Males.[1]

Make two score plot (using the timeplot function) of the combination
effect. In the first the PC1 scores of the Females and Males are plotted
as a function of the Groups. In the second the PC1 scores of the Groups
are plotted as a function of the Females and Males.[1]

3\. Make a loading plot for PC1 of the combination effect [1].

4\. What is your interpretation of the combination effect (Group +
Sex:Group) [1]

```{r Assignment 6G ASCA Combination model, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

```

#### Give your interpretation of 6G4 here

#### This concludes Assignment 6.

Make sure you save your .Rmd file and workspace so you can continue in
your next project session.

#### That concludes week 2 of the project!

Please hand in your .rmd and .pdf files on Canvas for your project
grade. The week 2 deadline is Tuesday January 20, 08:00 (AM).\*\*\*
